---
title: "Ozeki-Weedfish-2023"
output: html_document
date: "2025-01-15"
---

# Packages

A list of all the required packages

```{r}
required_packages <- c("ggplot2", "ggthemes", "ggfortify", "ggridges", "gghalves",
                       "ggExtra", "corrplot",# data visualisation 
                       
                      "performance", "see", "patchwork", "qqplotr", "glmmTMB", "DHARMa", # modelling 
                       
                       "datawizard", # data manipulation  
                       
                      "tidyverse", "janitor", "readxl", "broom", "data.table") # data tidy
```

A function to install and/or load the required package. This should report "All packages loaded" once it has finished loading them. Try running it twice if you do not see this message. 

```{r, message=FALSE, results='hide'}
loaded_packages <- lapply(required_packages, function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package)
    if (!require(package, character.only = TRUE)) {
      return(FALSE)
    }
  }
  return(TRUE)
})

#Check if all packages are loaded successfully
if (all(unlist(loaded_packages))) {
  cat("All packages loaded\n")
} else {
  cat("Some packages failed to load or install\n")
}
```

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(smatr)
library(RNOmni)
library(lme4)
library(car)
library(lmerTest)
library(merTools)
library(MuMIn)
library(emmeans)
library(corrplot)
library(ggthemes)
library(arm)
```

# Data wrangling

Here, I am creating a new data file where I am separating the fish ID components into multiple columns (batch, plant, housing tank, experiment tank). 

I am also making a relative total distance measure (total distance/time spent out of the refuge), making refuge a binary score (0 or 1), and adding the maximum time (2400 seconds) to "time_to_first_refuge" for fish that did not enter the refuge during their trial. 

```{r}
morph_raw <- read.csv("morph_raw.csv")
beh_raw <- read.csv("beh_raw.csv")
```


```{r}
# Creating new data file with the desired modifications
beh <- beh_raw %>%
  mutate(
    batch = substr(id, 2, 2),
    plant = if_else(nchar(id) == 15, substr(id, 11, 11), 
                    if_else(nchar(id) == 18, substr(id, 14, 14), NA_character_)),
    housing_tank = if_else(nchar(id) == 15, substr(id, 5, 5), 
                           if_else(nchar(id) == 18, substr(id, 8, 8), NA_character_)),
    experiment_tank = if_else(nchar(id) == 15, substr(id, 9, 9), 
                              if_else(nchar(id) == 18, substr(id, 12, 12), NA_character_)),
    total_distance_rel = total_distance/time_out_of_refuge,
    refuge = ifelse(is.na(time_to_first_refuge), 0, 1),
    time_to_first_refuge = if_else(is.na(time_to_first_refuge), 2400.000000, time_to_first_refuge),
    id = substr(id, 1, nchar(id) - 4)
  )

beh
```


```{r}
morph <- morph_raw %>% 
  rename_with(tolower) %>% 
  mutate(
    id = tolower(id),
    sex = tolower(sex)
  ) %>% 
  dplyr::select(-batch, -date) %>% 
  dplyr::mutate(treatment = ifelse(treatment == "control", "c",
                                   ifelse(treatment == "low", "l", "h")))
```

```{r}
# Extract the relevant part of the "id" column in "beh"
beh <- beh %>%
  mutate(
    matching_id = if_else(nchar(id) == 11, substr(id, 4, 7),
                          if_else(nchar(id) == 14, substr(id, 7, 10), NA_character_))
  ) 
```

```{r}
# Join the data frames based on the matching ID
merged_data <- dplyr::full_join(beh, morph, by = c("matching_id" = "id"))
```


```{r}
# Drop the unnecessary 'matching_id' column and keep only one 'treatment' column
wf <- merged_data %>%
  dplyr::select(-matching_id, -treatment.y) %>%  # Dropping the 'treatment.y' column
  dplyr::rename(treatment = treatment.x) %>% # Renaming 'treatment.x' to just 'treatment'
  filter(!is.na(id))
```

```{r}
view(wf)
```


## Blinding data

First creating a function for random allocation for treatment so we can do model selection blind.

```{r}
randomly_relabel <- function(data, column_names) {
  new_data <- data 
  for (col_name in column_names) {
    unique_factors <- unique(data[[col_name]])
    alphabet <- sample(letters, length(unique_factors), replace = TRUE)  # Change replace to TRUE
    new_labels <- setNames(alphabet, unique_factors)

# Use match to get the corresponding new label for each original factor
    new_data[[paste0(col_name, "_rand")]] <- new_labels[match(data[[col_name]], unique_factors)]
  }
  return(new_data)
}
```

Adding this new random allocation, it will change each time the code is run.

These columns will be names the same but with "_rand" at the end.

I have also removed the original columns so I don't see them when I open the data


```{r}
randomize_list <- c("id", "treatment")

beh_rand <- randomly_relabel(wf, randomize_list) %>% 
  dplyr::select(-any_of(randomize_list))

beh_rand
```
```{r}
beh_rand$housing_tank_fact <- as.factor(beh_rand$housing_tank)
beh_rand$experiment_tank_fact <- as.factor(beh_rand$experiment_tank)
```

# Behav

## Visualising raw beh data

```{r}
beh_rand %>% 
  ggplot(aes(x=time_to_first_refuge)) +
  geom_histogram() +
  coord_cartesian() +
  theme_few()

beh_rand %>% 
  ggplot(aes(x=time_out_of_refuge)) +
  geom_histogram() +
  coord_cartesian() +
  theme_few()

beh_rand %>% 
  ggplot(aes(x=total_distance)) +
  geom_histogram() +
  coord_cartesian() +
  theme_few()

beh_rand %>% 
  ggplot(aes(x=total_distance_rel)) +
  geom_histogram() +
  coord_cartesian() +
  theme_few()
```


```{r}
beh_rand %>% 
  dplyr::arrange(desc(total_distance)) %>% 
  dplyr::arrange(time_to_first_refuge)
```

## Building correlation matrix

```{r}
selected_columns <- beh_rand %>% 
  dplyr::select(time_to_first_refuge, time_out_of_refuge, total_distance, average_velocity, total_distance_rel)

correlation_matrix <- cor(selected_columns, use = "pairwise.complete.obs")
print(correlation_matrix)

corrplot(correlation_matrix)
```

# Total distance relative

```{r}
# Fit your original LMM 
total_distance_rel.lmm <- lmer(total_distance_rel ~ 
                     treatment_rand:sex 
                   + sex 
                   + treatment_rand 
                   + length
                   + (1|housing_tank_fact), data = beh_rand)
```

```{r}
check_model(total_distance_rel.lmm)
```

```{r}
check_normality(total_distance_rel.lmm)
```

### Model structure

```{r}
# Fit your original LMM 
total_distance_rel.lmm.log <- beh_rand %>% 
  dplyr::filter(!is.na(total_distance_rel)) %>% 
  lmer(log(total_distance_rel) ~ 
                     treatment_rand:sex 
                   + sex 
                   + treatment_rand 
                   + length
                   + (1|housing_tank_fact), data =.)
```

```{r}
beh_rand %>% 
  ggplot(aes(x = treatment_rand, y = total_distance_rel, fill = treatment_rand)) +
  geom_half_boxplot(side = "l", outlier.shape = NA, size = 1) +
  geom_half_violin(side = "r", alpha = 0.7, colour = "NA") +
  geom_half_point(side = "r", shape = 21) +
  scale_fill_manual(values=c("#F3B0C3", "#55CBCD", "#D0D2DF")) +
  theme_clean() + 
  theme(plot.subtitle = element_text(size = 13, face = "bold", colour = "deeppink"),
        legend.position = "none") +
  labs(subtitle = "Relative total distance", 
       caption = "Figure X. Relative total distance",
       x = "Treatment (random)",
       y = "Relative total distance")
```
### Checking assumptions

```{r}
check_model(total_distance_rel.lmm.log)
```

```{r}
check_predictions(total_distance_rel.lmm.log)
```

### Results
type III for interactions
type II for non interaction 
 
```{r}
signif(Anova(total_distance_rel.lmm.log, type = "III", test.statistic ="F"), digits = 3)
```

# Refuge use

56% of fish never entered refuge, we will use binary models

```{r}
# Calculate the proportion of fish that entered the refuge for each treatment
beh_rand %>% 
  dplyr::filter(!is.na(total_distance)) %>% 
  dplyr::group_by(treatment_rand) %>% 
  dplyr::reframe(n = length(refuge),
                 count = sum(refuge, na.rm = TRUE),
                 prop = count/n)
```

### Model structure 

```{r}
beh_rand <- beh_rand %>%
  mutate(length_standardized = scale(length))

refuge.glmm <- glmmTMB(refuge ~ 
                       treatment_rand:sex +
                       sex +
                       treatment_rand +
                       length_standardized +   # Use the standardized version of length
                       (1|housing_tank_fact), 
                       family = binomial, data = beh_rand)
```


### Check assumptions

```{r}
# Create simulated residuals
res_model <- simulateResiduals(fittedModel = refuge.glmm)

# Plot residuals
plot(res_model)
```
### Results

```{r}
summary(refuge.glmm)
```

```{r}
signif(Anova(refuge.glmm, type = "III"), digits = 3)
```
```{r}
# Create a function to reverse the randomization process
reverse_randomization <- function(data, random_columns) {
  # Initialize a copy of the data to modify
  new_data <- data
  
  # For each random column, get the corresponding original labels
  for (col in random_columns) {
    rand_column <- paste0(col, "_rand")
    
    # Get the unique random labels and their corresponding original labels
    random_labels <- unique(data[[rand_column]])
    original_labels <- unique(data[[col]]) 
    
    # Create a mapping from random labels to original labels
    label_mapping <- setNames(original_labels, random_labels)
    
    # Use the mapping to reverse the randomization
    new_data[[col]] <- label_mapping[data[[rand_column]]]
  }
  
  return(new_data)
}

# List of columns that were randomized
randomized_columns <- c("id", "treatment")

# Reverse the randomization for those columns
beh_original <- reverse_randomization(beh_rand, randomized_columns)

# View the data with the original treatment labels
head(beh_original)

```


